AFQ-Insight

### 2019-12-01 reports/manuscript/sections/conclusion.tex

We present here a novel method for analysis of dMRI tractometry data. This
method relies on the Sparse Group Lasso \cite{simon2013sparse} to provide both
accurate prediction of the phenotypic properties of individual subjects based on
their dMRI data, but also provides interpretable results by identifying the
features that are important across subjects to make these predictions. The
method is broadly applicable: it performs well in predicting both continuous
variables, such as biological age, as well as discrete variable, such as whether
a person is a patient or a healthy control. In both of these cases, SGL
out-performs previous algorithms that have been developed for these tasks. In
addition, the nested cross-validation approach that we have developed tunes the
degree of sparseness required by the algorithm, such that both very local
phenomena, as well as widely distributed phenomena are accurately captured.

Specifically, we demonstrated that the algorithm correctly detects the fact that
 ALS, which is a disease of lower motor neurons, is localized to the
 cortico-spinal tract. This recapitulates the results of previous analysis of
 these same data, using a targetted ROI-based approach. In contrast, in analysis
 of biological age, the coefficients identified by the algorithm are very widely
 distributed.

Taken together, these results demonstrate the promise of a group regularized
regression approach. Even at the scale of dozens of subjects, the results
provided by SGL are both accurate and interpretable. We expect that additional
data will improve the performance of these algorithms. Neuroscience has entered
an era in which large consortium efforts are putting together large datasets
that can be analyzed using this approach. Future work will apply this method to
these datasets. For example, to the Human Connectome Project dataset
\cite{VanEssen2012}

These results also motivate extensions of the method using more sophisticated
cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles. For example, a measurement could come from the corpus callosum, but
also from the right hemipshere. This would also require extending the cost
function used here to incorporate these constraints.

The method is packaged as open-source software called AFQ-Insight that is openly
available. The sofware integrates within a broader automated fiber
quantification software ecosystem: AFQ \cite{yeatman2012tract}, which extracts
tractometry data from raw dMRI, as well as AFQ-Browser, which visualizes
tractometry data and facilitates sharing of the results of dMRI studies
\cite{yeatman2018browser}.

@@

\begin{itemize}
  \item Feature selection + prediction
  \item Part of the AFQ ecosystem
  \item Open source, reproducibility
\end{itemize}

>>>>

We present here a novel method for analysis of dMRI tractometry data. This
method relies on the Sparse Group Lasso \cite{simon2013sparse} to provide both
accurate prediction of the phenotypic properties of individual subjects based on
their dMRI data, but also provides interpretable results by identifying the
features that are important across subjects to make these predictions. The
method is broadly applicable: it performs well in predicting both continuous
variables, such as biological age, as well as discrete variable, such as whether
a person is a patient or a healthy control. In both of these cases, SGL
out-performs previous algorithms that have been developed for these tasks
\cite{sarica2017corticospinal, Richard2018-ux}. The nested cross-validation
approach used to fit the model and make both predictioons and inferences from it
guards against overfitting and tunes the degree of sparseness required by the
algorithm, such that both phenomena that are locally focused in a particular
anatomical location and a particular property of diffusion (e.g., FA in the
CST), as well as widely distributed phenomena can be accurately captured.

Specifically, we demonstrated that the algorithm correctly detects the fact that
ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a targetted ROI-based approach
\cite{sarica2017corticospinal}. In contrast, in analysis of biological age, the
coefficients identified by the algorithm are very widely distributed in many
parts of the white matter, mirroring previous results with this dataset that
show a large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.

Taken together, these results demonstrate the promise of the group-regularized
regression approach. Even at the scale of dozens of subjects, the results
provided by SGL are both accurate and interpretable. Thus, this multivariate
analysis approach both (a) achieves high cross-validated accuracy for precision
medicine applications of dMRI data and (b) identifies relevant features of brain
anatomy that can further our neuroscientific understanding of clinical
disorders.

Meanwhile, neuroscience has entered an era in which consortium efforts are
putting together large datasets of high-quality dMRI data to address a variety
of scientific questions \cite{jernigan2016ping, jernigan2018abcd,
alexander2017open, Miller2016-hw, VanEssen2012}. The volume and complexity of
these data pose a substantial challenge. Dimensionality reduction with
tractometry, followed by analysis with the approach we present here promises to
capitalize on the wealth of data and on the co-measurement of interesting and
important phenotypical data about brain health and about the participants'
cognitive abilities. We also expect the group-regularized approach to improve
with larger datasets.

@@

The results we present here also motivate extensions of the method using more
sophisticated cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles \ariel{citation needed}. For example, a measurement could come from the
corpus callosum, but also from the right hemipshere. This would also require
extending the cost function used here to incorporate these constraints.

The method is packaged as open-source software called AFQ-Insight that is openly
available, and this also where extensions of the method will be developed. The
sofware integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract}, which extracts tractometry data from raw
and processed dMRI datasets, as well as AFQ-Browser, which visualizes
tractometry data and facilitates sharing of the results of dMRI studies
\cite{yeatman2018browser}. To facilitate reproducibility and ease use of the
software, the results presented in this paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks
\cite{kluyver2016jupyter}

### 2019-12-01 reports/manuscript/sections/introduction.tex

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001-uz}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., ). Moreover, use of SGL in particular helps derive informative
properties of the human white matter, both in cases where the underlying causes
of variance are sparsely concentrated (e.g., in one particular white matter
tract) and in cases where biological substrates of differences between
individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to  which and that can be set empirically,
based on the data, in a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

>>>>

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001-uz}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., whether an individual has a particular disease). Moreover, use of
SGL in particular helps derive informative properties of the human white matter,
both in cases where the underlying causes of variance are sparsely concentrated
(e.g., in one particular white matter tract) and in cases where biological
substrates of differences between individuals are more widespread. This is
because SGL is controlled by meta-parameters that set the degree to  which and
that can be set empirically, based on the data, in a process of nested
cross-validation. We demonstrate these properties of the algorithm in data from
previously published studies \cite{sarica2017corticospinal,
yeatman2014lifespan}.
